---
title: "R Notebook"
output: html_notebook
---

Linear Mixed Models
Adapted from: https://chenzixu.rbind.io/slides/lme/lmer.html


```{r}
library(lme4)
library(magrittr)
library(tidyverse)
library(ggplot2)
library(lmerTest)
```

```{r}
data("sleepstudy")
head(sleepstudy)
```

Plotting subjectwise data

```{r}
sleep2 <- sleepstudy %>%
  filter(Days >= 2) %>%
  mutate(days_deprived = Days - 2)
ggplot(sleep2, aes(x = days_deprived, 
                   y = Reaction)) +
  geom_point() +
  scale_x_continuous(breaks = 0:7) +
  facet_wrap(~Subject) +
  labs(y = "Reaction Time", 
       x = "Days deprived of sleep (0 = baseline)")
```


Complete Pooling model


```{r}
cp_model <- lm(Reaction ~ days_deprived, sleep2)
summary(cp_model)
ggplot(sleep2, aes(x = days_deprived, y = Reaction)) +
  geom_abline(intercept = coef(cp_model)[1],
              slope = coef(cp_model)[2],
              color = '#f4cae2', size = 1.5) +
  geom_point() +
  scale_x_continuous(breaks = 0:7) +
  facet_wrap(~Subject, nrow = 3) +
  labs(y = "Reaction Time", 
       x = "Days deprived of sleep (0 = baseline)")
```

No Pooling Model

```{r}
sleep2 %>% pull(Subject) %>% is.factor()
np_model <- lm(Reaction ~ days_deprived + Subject + days_deprived:Subject,
               data = sleep2)
  summary(np_model)
all_intercepts <- c(coef(np_model)["(Intercept)"],
                    coef(np_model)[3:19] + coef(np_model)["(Intercept)"])
all_slopes  <- c(coef(np_model)["days_deprived"],
                 coef(np_model)[20:36] + coef(np_model)["days_deprived"])
ids <- sleep2 %>% pull(Subject) %>% levels() %>% factor()
np_coef <- tibble(Subject = ids,
                  intercept = all_intercepts,
                  slope = all_slopes)
```

```{r}
ggplot(sleep2, aes(x = days_deprived, y = Reaction)) +
  geom_abline(data = np_coef,
              mapping = aes(intercept = intercept,
                            slope = slope),
              color = '#f4cae2', size = 1.5) +
  geom_point() + theme_bw() +
  scale_x_continuous(breaks = 0:7) +
  facet_wrap(~Subject, nrow=3) +
  labs(y = "Reaction Time", 
       x = "Days deprived of sleep (0 = baseline)")
```

Partial Pooling Model

```{r}
pp_mod <- lmer(Reaction ~ days_deprived + (days_deprived | Subject), sleep2)
summary(pp_mod)
newdata <- crossing(
  Subject = sleep2 %>% pull(Subject) %>% levels() %>% factor(),
  days_deprived = 0:7)
newdata2 <- newdata %>%
  mutate(Reaction = predict(pp_mod, newdata))
```


```{r}
ggplot(sleep2, aes(x = days_deprived, y = Reaction)) +
  geom_line(data = newdata2,
            color = '#f4cae2', size = 1.5) +
  geom_point() + theme_bw() +
  scale_x_continuous(breaks = 0:7) +
  facet_wrap(~Subject, nrow = 3) +
  labs(y = "Reaction Time", 
       x = "Days deprived of sleep (0 = baseline)")
```

##Research Question
We're interested in the relationship between pitch and politeness (Winter & Grawunder, 2012).

Politeness: formal/polite and informal register (categorical factor)
multiple measures per subject (inter-dependent!)

```{r}
data = read.csv("http://www.bodowinter.com/tutorial/politeness_data.csv")
head(data)
```

Convert attitude, gender, subject into factors.
```{r}
data = data %>% mutate(attitude=as.factor(attitude), gender=as.factor(gender), subject=as.factor(subject))

```


Random Intercept Models

```{r}
politeness.model0 = lmer(frequency ~ attitude + (1|subject) + (1|scenario), data=data)
summary(politeness.model0)
```

```{r}
politeness.model = lmer(frequency ~ attitude + gender + (1|subject) + (1|scenario), data=data)
summary(politeness.model)
```
Likelihood Ratio Test

```{r}
politeness.null = lmer(frequency ~ gender + (1|subject) + (1|scenario), data=data, REML=FALSE)
politeness.full = lmer(frequency ~ attitude + gender + (1|subject) + (1|scenario), data=data, REML=FALSE)
anova(politeness.null, politeness.full)
```

Random Slope Model

```{r}
politeness.model1 = lmer(frequency~attitude + gender + (1+attitude|subject) + (1+attitude|scenario), data = data)
coef(politeness.model1)
```


