---
title: "Regression"
output: html_document
date: "2022-11-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path)) # set current path as work path
```



```{r}
load( file.path("parenthood.Rdata" ))
colnames(parenthood) <- c('dan.sleep','baby.sleep','dan.grump','day')
```


```{r}
head(parenthood)
```

## Simple linear regression:
dan.grump ~ dan.sleep (Is Dan's sleep deprivation leading to grumpiness the next day?)
```{r}
regression.1 <- lm( formula = dan.grump ~ dan.sleep,  
                    data = parenthood )
print( regression.1 )
```

```{r}
summary(regression.1)
```

```{r}
cor.test( x = parenthood$dan.sleep, y = parenthood$dan.grump )
```


## Multiple linear regression
dan.grump ~ dan.sleep + baby.sleep (Is Dan's sleep and the baby's sleep both together leading to grumpiness the next day?)

```{r}
regression.2 <- lm( formula = dan.grump ~ dan.sleep + baby.sleep,  
                     data = parenthood )
print( regression.2 )
```
```{r}
summary(regression.2)
```

```{r}
library(lsr)
correlate(parenthood, test=TRUE)
```

```{r}
confint( object = regression.2,
        level = .99)
```

```{r}
standardCoefs( regression.2 )

```

## Checking Model Assumptions
First, we need to generate the residuals!

```{r}
residuals( object = regression.2 )
```

```{r}
rstandard( model = regression.2 )
```

```{r}
rstudent( model = regression.2 )
```

```{r}
hatvalues( model = regression.2 )
```

```{r}
ckd <- cooks.distance( model = regression.2 )
ckd
```

Directly plotting Cook's distance using plot from car package
```{r}
plot(regression.2,which=4)
```




```{r}
lm( formula = dan.grump ~ dan.sleep + baby.sleep,  # same formula
     data = parenthood,       # same data frame...
     subset = -64             # ...but observation 64 is deleted
 )
```




```{r}
 hist( x = residuals( regression.2 ),   # data are the residuals
       xlab = "Value of residual",      # x-axis label
       main = "",                       # no title 
       breaks = 20                      # lots of breaks
 )
```


Directly make the QQ plot of residuals

```{r}
plot( x = regression.2, which = 2 )
```













Backward elimination

```{r}
full.model <- lm( formula = dan.grump ~ dan.sleep + baby.sleep + day,  
                   data = parenthood  
 )
```

```{r}
 step( object = full.model,     # start at the full model
       direction = "backward"   # allow it remove predictors but not add them
 )
```
Forward selection
```{r}
 null.model <- lm( dan.grump ~ 1, parenthood )   # intercept only.
 step( object = null.model,     # start with null.model
       direction = "forward",   # only consider "addition" moves
       scope =  dan.grump ~ dan.sleep + baby.sleep + day  # largest model allowed
 )
```
 Comparing two regression models
```{r}
M0 <- lm( dan.grump ~ dan.sleep + day, parenthood )
M1 <- lm( dan.grump ~ dan.sleep + day + baby.sleep, parenthood )
```

```{r}
AIC( M0, M1 )
```

```{r}
anova( M0, M1 )
```

